\bf Properties and Bayesian fitting of restricted Boltzmann machines

### 2018-09-01 writing/paper-rbm/docs/paper.tex
trying out some things

\documentclass[12pt]{article}
\usepackage{amsmath,amssymb}
\usepackage{graphicx,psfrag,epsf}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{url} % not crucial - just used below for the URL

\usepackage{longtable,booktabs}
\usepackage[unicode=true]{hyperref}
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

@@

\date{}
\maketitle
\begin{abstract}
A restricted Boltzmann machine (RBM) is an undirected graphical model
constructed for discrete or continuous random variables, with two
layers, one hidden and one visible, and no conditional dependency within
a layer. In recent years, RBMs have risen to prominence due to their
connection to deep learning. By treating a hidden layer of one RBM as
the visible layer in a second RBM, a deep architecture can be created.
RBMs are thought to thereby have the ability to encode very complex and
rich structures in data, making them attractive for supervised learning.
However, the generative behavior of RBMs is largely unexplored. In this
paper, we discuss the relationship between RBM parameter specification
in the binary case and model properties such as degeneracy, instability
and uninterpretability. We also describe the difficulties that arise in
likelihood-based and Bayes fitting of such (highly flexible) models,
especially as Gibbs sampling (quasi-Bayes) methods are often advocated
for the RBM model structure.
\end{abstract}
\noindent%
{\it Keywords:}  Degeneracy, Instability, Classification, Deep Learning, Graphical Models
\vfill

\newpage
\spacingset{1.45}

@@

The data mining and machine learning communities have recently shown
great interest in deep learning, specifically in stacked restricted
Boltzmann machines (RBMs) (see R. Salakhutdinov and Hinton
\protect\hyperlink{ref-salakhutdinov2009deep}{2009},
\protect\hyperlink{ref-salakhutdinov2012efficient}{2012}; Srivastava,
Salakhutdinov, and Hinton
\protect\hyperlink{ref-srivastava2013modeling}{2013}; Le Roux and Bengio
\protect\hyperlink{ref-le2008representational}{2008} for examples). A
RBM is a probabilistic undirected graphical model (for discrete or
continuous random variables) with two layers, one hidden and one
visible, with no conditional dependency within a layer (Smolensky
\protect\hyperlink{ref-smolensky1986information}{1986}). These models
have reportedly been used with success in classification of images
(Larochelle and Bengio
\protect\hyperlink{ref-larochelle2008classification}{2008}; Srivastava
and Salakhutdinov
\protect\hyperlink{ref-srivastava2012multimodal}{2012}). However, the
model properties are largely unexplored in the literature and the
commonly cited fitting methodology remains heuristic-based and relies on
rough approximation (Hinton, Osindero, and Teh
\protect\hyperlink{ref-hinton2006fast}{2006}). In this paper, we provide
steps toward a fuller understanding of the model class and its
properties from the perspective of statistical model theory, and we then
explore the possibility of a rigorous fitting methodology. We find the
RBM model class to be concerning in two fundamental ways.

@@

In addition to model properties, we also explore the quality of
estimation in fitting these models. The fitting can be problematic for
two reasons, the first being computational and the second concerning
flexibility. As the size of these models grows, both maximum likelihood
and Bayesian methods of fitting quickly become intractable. The
literature often suggests Markov chain Monte Carlo (MCMC) tools for
approximate maximization of likelihoods to fit these models (e.g., Gibbs
sampling to exploit conditional structure in hidden and visible
variables), but little is said about the attributes of realized
estimates (Hinton \protect\hyperlink{ref-hinton2010practical}{2010};
Hinton, Osindero, and Teh \protect\hyperlink{ref-hinton2006fast}{2006}).
Related to this, these MCMC algorithms require updating potentially many
latent variables (hiddens) which can critically influence convergence in
MCMC-based likelihood methods. Applying basic statistical principles in
fitting RBM models of tractable sizes, we compare three fully Bayesian
techniques involving MCMC, which are computationally more accessible
than direct maximum likelihood and also aim to avoid parts of a RBM
parameter space that yield unattractive models. As might be expected,
with greater computational complexity comes an increase in fitting
accuracy, but at the cost of practical feasibility.

@@

This paper is structured as follows. Section \ref{background} formally
defines the RBM including the joint distribution of hidden and visible
variables and explains the model's connection to deep learning.
Additionally, measures of model impropriety and methods of
quantifying/detecting it are defined. Section
\ref{explorations-of-model-properties-through-simulation} details our
explorations into the model behavior and potential propriety issues with
the RBM class. We examine three Bayesian fitting techniques intended to
avoid model impropriety issues raised in Section \ref{model-fitting} and
conclude with a discussion in Section \ref{discussion}. Supplementary
online material provides proofs for results on RBM parameterizations and
data codings described in Section \ref{data-encoding}.

@@

Let \(\boldsymbol x = (h_1, \dots, h_\nh, v_1,\dots,v_\nv)\) represent
the states of the visible and hidden nodes in a RBM for some integers
\(\nv, \nh \ge 1\). Each single binary random variable, visible or
hidden, will take its values in a common coding set \(\mathcal{C}\),
where we allow one of two possibilities for the coding set,
\(\mathcal{C}=\{0,1\}\) or \(\mathcal{C}=\{-1,1\}\), with ``\(1\)''
always indicating the ``high'' value of the variable. While
\(\mathcal{C}=\{0,1\}\) may be a natural starting point, we argue in
Section \ref{explorations-of-model-properties-through-simulation} that
the coding \(\mathcal{C}=\{-1,1\}\) induces more interpretable model
properties for the RBM. A standard parametric form for probabilities
corresponding to a potential vector of states,
\(X = (H_1, \dots, H_\nh, V_1,\dots,V_\nv)\), for the nodes is
\begin{align}
\label{eq:pmf}
f_{\boldsymbol \theta} (\boldsymbol x)\equiv P_{\boldsymbol \theta}(\boldsymbol X = \boldsymbol x) = \frac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\gamma(\boldsymbol \theta)}, \quad \boldsymbol x \in \mathcal{C}^{\nh + \nv} 
\end{align} where
\(\boldsymbol \theta \equiv (\theta_{11}, \dots, \theta_{1\nh}, \dots, \theta_{\nv 1}, \dots, \theta_{\nv \nh}, \theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}) \in \mathbb{R}^{\nv + \nh + \nv*\nh}\)
denotes the vector of model parameters and the denominator \[
\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{C}^{\nh+\nv}}\exp\left(\sum_{i = 1}^\nv \sum_{j=1}^\nh \theta_{ij} v_i h_j + \sum_{i = 1}^\nv\theta_{v_i} v_i + \sum_{j = 1}^\nh\theta_{h_j} h_j\right)
\] is the normalizing function that ensures the probabilities
(\eqref{eq:pmf}) sum to one. For
\(\boldsymbol x = (h_1, \dots, h_\nh, v_1, \dots, v_\nh) \in \mathcal{C}^{\nv + \nh}\)
and \begin{align}
\boldsymbol t(\boldsymbol x) &= (h_1, \dots, h_\nh, v_1, \dots, v_\nv, v_1 h_1, \dots, v_\nv h_\nh) \in \mathcal{C}^{\nh + \nv + \nh*\nv}, \label{eq:t}
\end{align} let
\(\mathcal{T} = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nv + \nh}\} \subset \mathbb{R}^{\nv + \nh + \nv * \nh}\)
be the set of possible values for the vector of variables needed to
compute probabilities (\eqref{eq:pmf}) in the model, and write
\(Q_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{i = 1}^\nh\sum\limits_{j=1}^\nv \theta_{ij} h_i v_j + \sum\limits_{i = 1}^\nh\theta_{hi} h_i + \sum\limits_{j = 1}^\nv\theta_{vj} v_j\)
for the ``neg-potential'' function. The RBM model is parameterized by
\(\boldsymbol \theta\) containing two types of parameters, main effects
and interaction effects. The main effects parameters
(\(\{\theta_{v_i}, \theta_{h_j}\}_{\substack{i = 1, \dots, \nv,\\j = 1, \dots, \nh}}\))
weight the values of the visible \(v_i\) and hidden \(h_j\) nodes in
probabilities (\eqref{eq:pmf}) and the interaction effect parameters
(\(\theta_{ij}\)) weight the values of the connections \(v_i h_j\), or
dependencies, between hidden and visible layers.

@@

RBMs have risen to prominence in recent years due to their connection to
deep learning (see Hinton, Osindero, and Teh
\protect\hyperlink{ref-hinton2006fast}{2006}; R. Salakhutdinov and
Hinton \protect\hyperlink{ref-salakhutdinov2012efficient}{2012};
Srivastava, Salakhutdinov, and Hinton
\protect\hyperlink{ref-srivastava2013modeling}{2013} for examples). By
stacking multiple layers of RBMs in a deep architecture, proponents of
the model are claimed to produce the ability to learn ``internal
representations that become increasingly complex, which is considered to
be a promising way of solving object and speech recognition problems''
(R. Salakhutdinov and Hinton
\protect\hyperlink{ref-salakhutdinov2009deep}{2009}, 450). The stacking
is achieved by treating a hidden layer of one RBM as the visible layer
in a second RBM, and so on, until the desired multi-layer architecture
is created.

@@

In Random Graph Model theory, \emph{model degeneracy} means there is a
disproportionate amount of probability placed on only a few elements of
the sample space, \(\mathcal{X}\), by the model (Handcock
\protect\hyperlink{ref-handcock2003assessing}{2003}). For random graph
models, \(\mathcal{X}\) denotes all possible graphs that can be
constructed from a set of nodes and an exponentially parameterized
random graph model has a distribution with a probability mass function
of the form \begin{align*}
f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)}{\gamma(\boldsymbol \theta)}, \boldsymbol x \in \mathcal{X},
\end{align*} where
\(\boldsymbol \theta \in \Theta \subset \mathbb{R}^q\) is the model
parameter, and \(\boldsymbol t: \mathcal{X} \rightarrow \mathbb{R}^q\)
is a vector of statistics based on the adjacency matrix of a graph.
Here, as earlier,
\(\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)\)
is the normalizing function. Let \(C\) denote the convex hull of the
potential outcomes of sufficient statistics,
\(\{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}\),
under the model above. Handcock
(\protect\hyperlink{ref-handcock2003assessing}{2003}) classifies an
exponentially parametrized random graph model at \(\boldsymbol \theta\)
as \emph{near-degenerate} if the mean value of the vector of sufficient
statistics under \(\boldsymbol \theta\),
\(\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)\),
is close to the boundary of \(C\). Intuitively, if a model is
near-degenerate in the sense that only a small number of elements of the
sample space \(\mathcal{X}\) have positive probability, the expected
value \(\text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)\) is
an average of that same small number of values of \(t( \boldsymbol x)\)
(defining the boundary of the hull \(C\)) and can be expected to
\emph{not} be pulled deep into the interior of \(C\).

A RBM model can be thought to exhibit an analogous form of
\emph{near-degeneracy} when there is a disproportionate amount of
probability placed on a small number of elements in the sample space of
visibles and hiddens, \(\mathcal{C}^{\nv + \nh}\). Using the idea of
Handcock (\protect\hyperlink{ref-handcock2003assessing}{2003}), when the
random vector
\(\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_\nv, h_1, \dots, h_\nh, v_1 h_1, \dots, v_V h_\nh ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nh + \nv} \}\)
from (\eqref{eq:t}), appearing in the neg-potential function
\(Q_{\boldsymbol \theta}(\cdot)\), has a mean vector
\(\boldsymbol \mu(\boldsymbol \theta) \in \mathbb{R}^{\nv + \nh + \nv * \nh}\)
close to the boundary of the convex hull of \(\mathcal{T}\), then the
RBM model can be said to exhibit near-degeneracy at
\(\boldsymbol \theta \in \mathbb{R}^{\nv + \nh + \nh*\nv}\). Here the
mean of \(\boldsymbol t(\boldsymbol x)\) is \begin{align*}
\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol X)  &= \sum\limits_{x \in \mathcal{C}^{\nv + \nh}} \left\{ \boldsymbol t(\boldsymbol x) f_{\boldsymbol \theta} (\boldsymbol x) \right\} \\
&= \sum\limits_{x \in \mathcal{C}^{\nv + \nh}} \left\{ \boldsymbol t(\boldsymbol x)\dfrac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}}\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}\right\}.
\end{align*}
\hypertarget{instability}{%
\subsubsection{Instability}\label{instability}}

@@

In formulating a RBM model for a potentially large number of visibles
(i.e., as \(\nv \rightarrow \infty\)), we will say that the ratio
(\eqref{eq:elpr}) needs to stay bounded for a sequence of RBM models to be
stable. That is, we make the following convention.
\BeginKnitrBlock{definition}[S-unstable RBM]
\protect\hypertarget{def:unnamed-chunk-1}{}{\label{def:unnamed-chunk-1}
\iffalse (S-unstable RBM) \fi{} }Let
\(\boldsymbol \theta_\nv \in \mathbb{R}^{\nv + \nh + \nh*\nv}, \nv \ge 1\),
be an element of a sequence of RBM parameters where the number of
hiddens, \(\nh \equiv \nh(\nv) \ge 1\), can be an arbitrary function of
the number \(\nv\) of visibles. A RBM model formulation is
\emph{Schweinberger-unstable} or \emph{S-unstable} if \begin{align*}
\lim\limits_{\nv \rightarrow \infty} \frac{1}{\nv} \text{LREP}(\boldsymbol \theta_\nv) = \infty.
\end{align*}
\EndKnitrBlock{definition}

@@

\BeginKnitrBlock{proposition}
\protect\hypertarget{prp:instab1}{}{\label{prp:instab1} }Let \(c > 0\) and
let \(\text{LREP}(\boldsymbol \theta_\nv)\) be as in (\eqref{eq:elpr}) for
an integer \(\nv \ge 1\). If
\(\frac{1}{\nv}\text{LREP}(\boldsymbol \theta_\nv) > c\), then
\(\Delta(\boldsymbol \theta_\nv) > c\).
\EndKnitrBlock{proposition}

>>>>

\documentclass[]{article}
%\usepackage{nips15submit_e,times}
\usepackage{url}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{longtable,booktabs}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Properties and Bayesian fitting of restricted Boltzmann machines},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{5}

% Tech stuff starts here
\usepackage{psfrag,epsf}
%\pdfminorversion=4

@@

\maketitle
\begin{abstract}
A restricted Boltzmann machine (RBM) is an undirected graphical model
constructed for discrete or continuous random variables, with two
layers, one hidden and one visible, and no conditional dependency within
a layer. In recent years, RBMs have risen to prominence due to their
connection to deep learning. By treating a hidden layer of one RBM as
the visible layer in a second RBM, a deep architecture can be created.
RBMs are thought to thereby have the ability to encode very complex and
rich structures in data, making them attractive for supervised learning.
However, the generative behavior of RBMs is largely unexplored and
typical fitting methodology does not easily allow for uncertainty
quantification in addition to point estimates. In this paper, we discuss
the relationship between RBM parameter specification in the binary case
and model properties such as degeneracy, instability and
uninterpretability. We also describe the difficulties that arise in
likelihood-based and discuss potential Bayes fitting of such (highly
flexible) models, especially as Gibbs sampling (quasi-Bayes) methods are
often advocated for the RBM model structure.
\end{abstract}
\noindent%
{\it Keywords:}  Degeneracy, Instability, Classification, Deep Learning, Graphical Models
\vfill

\hfill {\tiny technometrics tex template (do not remove)}
\newpage
\spacingset{1.45} % DON'T change the spacing!

@@

The data mining and machine learning communities have recently shown
great interest in deep learning, specifically in stacked restricted
Boltzmann machines (RBMs) (see Salakhutdinov and Hinton
\protect\hyperlink{ref-salakhutdinov2009deep}{2009},
\protect\hyperlink{ref-salakhutdinov2012efficient}{2012}; Srivastava,
Salakhutdinov, and Hinton
\protect\hyperlink{ref-srivastava2013modeling}{2013}; Le Roux and Bengio
\protect\hyperlink{ref-le2008representational}{2008} for examples). A
RBM is a probabilistic undirected graphical model (for discrete or
continuous random variables) with two layers, one hidden and one
visible, with no conditional dependency within a layer (Smolensky
\protect\hyperlink{ref-smolensky1986information}{1986}). These models
have reportedly been used with success in classification of images
(Larochelle and Bengio
\protect\hyperlink{ref-larochelle2008classification}{2008}; Srivastava
and Salakhutdinov
\protect\hyperlink{ref-srivastava2012multimodal}{2012}). However, the
model properties are largely unexplored in the literature and the
commonly cited fitting methodology remains heuristic-based and relies on
rough approximation (Hinton, Osindero, and Teh
\protect\hyperlink{ref-hinton2006fast}{2006}). Additionally, the current
fitting methodology does not easily allow for interval estimation of
parameters in the model, and instead only provides for point estimation.
In this paper, we provide steps toward a fuller understanding of the
model class and its properties from the perspective of statistical model
theory, and we then explore the possibility of a rigorous fitting
methodology with natural uncertainty quantification. We find the RBM
model class to be concerning in two fundamental ways.

@@

Separate of the exploration of model properties, we also investigate the
quality of estimation in fitting these models and propose an estimation
procedure that will allow for uncertainty quantification rather than
point estimates alone. The fitting of RBMs can be problematic for two
reasons, the first being computational and the second concerning
flexibility. As the size of these models grows, both maximum likelihood
and Bayesian methods of fitting quickly become intractable. The
literature often suggests Markov chain Monte Carlo (MCMC) tools for
approximate maximization of likelihoods to fit these models (e.g., Gibbs
sampling to exploit conditional structure in hidden and visible
variables), but little is said about the attributes of realized
estimates (Hinton \protect\hyperlink{ref-hinton2010practical}{2010};
Hinton, Osindero, and Teh \protect\hyperlink{ref-hinton2006fast}{2006}),
including the absence of interval estimates.

Furthermore, these MCMC algorithms require updating potentially many
latent variables (hiddens) which can critically influence convergence in
MCMC-based likelihood methods. Applying basic statistical principles in
fitting RBM models of tractable sizes, we compare three fully Bayesian
techniques involving MCMC, which are computationally more accessible
than direct maximum likelihood and also aim to avoid parts of a RBM
parameter space that yield unattractive models. As might be expected,
with greater computational complexity comes an increase in fitting
accuracy, but at the cost of practical feasibility.

@@

This paper is structured as follows. Section \ref{background} formally
defines the RBM including the joint distribution of hidden and visible
variables and explains the model's connection to deep learning.
Additionally, measures of model impropriety and methods of
quantifying/detecting it are defined. Section
\ref{explorations-of-model-properties-through-simulation} details our
explorations into the model behavior and potential propriety issues with
the RBM class. This follows with the call to action in Sculley et al.
(\protect\hyperlink{ref-sculley2018winner}{2018}) to employ small toy
examples to advance empirical understanding of deep learning models and
strive for a ``higher level of empirical rigor in the field.'' We
examine three Bayesian fitting techniques intended to avoid model
impropriety issues raised in Section \ref{model-fitting} and conclude
with a discussion in Section \ref{discussion}. Supplementary online
material provides proofs for results on RBM parameterizations and data
codings described in Section \ref{data-encoding}.

@@

Let \(\boldsymbol x = (h_1, \dots, h_\nh, v_1,\dots,v_\nv)\) represent
the states of the visible and hidden nodes in a RBM for some integers
\(\nv, \nh \ge 1\). Each single binary random variable, visible or
hidden, will take its values in a common coding set \(\mathcal{C}\),
where we allow one of two possibilities for the coding set,
\(\mathcal{C}=\{0,1\}\) or \(\mathcal{C}=\{-1,1\}\), with ``\(1\)''
always indicating the ``high'' value of the variable. While
\(\mathcal{C}=\{0,1\}\) may be a natural starting point, we argue in
Section \ref{explorations-of-model-properties-through-simulation} that
the coding \(\mathcal{C}=\{-1,1\}\) induces more interpretable model
properties for the RBM. A standard parametric form for probabilities
corresponding to a potential vector of states,
\(X = (H_1, \dots, H_\nh, V_1,\dots,V_\nv)\), for the nodes is
\begin{align}
\label{eq:pmf}
f_{\boldsymbol \theta} (\boldsymbol x)\equiv P_{\boldsymbol \theta}(\boldsymbol X = \boldsymbol x) = \frac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\gamma(\boldsymbol \theta)}, \quad \boldsymbol x \in \mathcal{C}^{\nh + \nv} 
\end{align} where
\(\boldsymbol \theta \equiv (\theta_{11}, \dots, \theta_{1\nh}, \dots, \theta_{\nv 1}, \dots, \theta_{\nv \nh}, \theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}) \in \mathbb{R}^{\nv + \nh + \nv*\nh}\)
denotes the vector of model parameters and the denominator \[
\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{C}^{\nh+\nv}}\exp\left(\sum_{i = 1}^\nv \sum_{j=1}^\nh \theta_{ij} v_i h_j + \sum_{i = 1}^\nv\theta_{v_i} v_i + \sum_{j = 1}^\nh\theta_{h_j} h_j\right)
\] is the normalizing function that ensures the probabilities
\eqref{eq:pmf} sum to one. For
\(\boldsymbol x = (h_1, \dots, h_\nh, v_1, \dots, v_\nh) \in \mathcal{C}^{\nv + \nh}\)
and \begin{align}
\boldsymbol t(\boldsymbol x) &= (h_1, \dots, h_\nh, v_1, \dots, v_\nv, v_1 h_1, \dots, v_\nv h_\nh) \in \mathcal{C}^{\nh + \nv + \nh*\nv}, \label{eq:t}
\end{align} let
\(\mathcal{T} = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nv + \nh}\} \subset \mathbb{R}^{\nv + \nh + \nv * \nh}\)
be the set of possible values for the vector of variables needed to
compute probabilities \eqref{eq:pmf} in the model, and write
\(Q_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{i = 1}^\nh\sum\limits_{j=1}^\nv \theta_{ij} h_i v_j + \sum\limits_{i = 1}^\nh\theta_{hi} h_i + \sum\limits_{j = 1}^\nv\theta_{vj} v_j\)
for the ``neg-potential'' function. The RBM model is parameterized by
\(\boldsymbol \theta\) containing two types of parameters, main effects
and interaction effects. The main effects parameters
(\(\{\theta_{v_i}, \theta_{h_j}\}_{\substack{i = 1, \dots, \nv,\\j = 1, \dots, \nh}}\))
weight the values of the visible \(v_i\) and hidden \(h_j\) nodes in
probabilities \eqref{eq:pmf} and the interaction effect parameters
(\(\theta_{ij}\)) weight the values of the connections \(v_i h_j\), or
dependencies, between hidden and visible layers.

@@

RBMs have risen to prominence in recent years due to their connection to
deep learning (see Hinton, Osindero, and Teh
\protect\hyperlink{ref-hinton2006fast}{2006}; Salakhutdinov and Hinton
\protect\hyperlink{ref-salakhutdinov2012efficient}{2012}; Srivastava,
Salakhutdinov, and Hinton
\protect\hyperlink{ref-srivastava2013modeling}{2013} for examples). By
stacking multiple layers of RBMs in a deep architecture, proponents of
the model are claimed to produce the ability to learn ``internal
representations that become increasingly complex, which is considered to
be a promising way of solving object and speech recognition problems''
(Salakhutdinov and Hinton
\protect\hyperlink{ref-salakhutdinov2009deep}{2009}, 450). The stacking
is achieved by treating a hidden layer of one RBM as the visible layer
in a second RBM, and so on, until the desired multi-layer architecture
is created.

@@

In Random Graph Model theory, \emph{model degeneracy} means there is a
disproportionate amount of probability placed on only a few elements of
the sample space, \(\mathcal{X}\), by the model (Handcock
\protect\hyperlink{ref-handcock2003assessing}{2003}). For random graph
models, \(\mathcal{X}\) denotes all possible graphs that can be
constructed from a set of nodes and an exponentially parameterized
random graph model has a distribution with a probability mass function
of the form \begin{align*}
f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)}{\gamma(\boldsymbol \theta)}, \boldsymbol x \in \mathcal{X},
\end{align*} where
\(\boldsymbol \theta \in \Theta \subset \mathbb{R}^q\) is the model
parameter, and \(\boldsymbol t: \mathcal{X} \rightarrow \mathbb{R}^q\)
is a vector of statistics based on the adjacency matrix of a graph.
Here, as earlier,
\(\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)\)
is the normalizing function. Let \(C\) denote the convex hull of the
potential outcomes of sufficient statistics,
\(\{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}\),
under the model above. Handcock
(\protect\hyperlink{ref-handcock2003assessing}{2003}) classifies an
exponentially parameterized random graph model at \(\boldsymbol \theta\)
as \emph{near-degenerate} if the mean value of the vector of sufficient
statistics under \(\boldsymbol \theta\),
\(\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)\),
is close to the boundary of \(C\). Intuitively, if a model is
near-degenerate in the sense that only a small number of elements of the
sample space \(\mathcal{X}\) have positive probability, the expected
value \(\text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)\) is
an average of that same small number of values of \(t( \boldsymbol x)\)
(defining the boundary of the hull \(C\)) and can be expected to
\emph{not} be pulled deep into the interior of \(C\).

A RBM model can be thought to exhibit an analogous form of
\emph{near-degeneracy} when there is a disproportionate amount of
probability placed on a small number of elements in the sample space of
visibles and hiddens, \(\mathcal{C}^{\nv + \nh}\). Using the idea of
Handcock (\protect\hyperlink{ref-handcock2003assessing}{2003}), when the
random vector
\(\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_\nv, h_1, \dots, h_\nh, v_1 h_1, \dots, v_V h_\nh ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nh + \nv} \}\)
from \eqref{eq:t}, appearing in the neg-potential function
\(Q_{\boldsymbol \theta}(\cdot)\), has a mean vector
\(\boldsymbol \mu(\boldsymbol \theta) \in \mathbb{R}^{\nv + \nh + \nv * \nh}\)
close to the boundary of the convex hull of \(\mathcal{T}\), then the
RBM model can be said to exhibit near-degeneracy at
\(\boldsymbol \theta \in \mathbb{R}^{\nv + \nh + \nh*\nv}\). Here the
mean of \(\boldsymbol t(\boldsymbol x)\) is \begin{align*}
\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol X)  &= \sum\limits_{x \in \mathcal{C}^{\nv + \nh}} \left\{ \boldsymbol t(\boldsymbol x) f_{\boldsymbol \theta} (\boldsymbol x) \right\} \\
&= \sum\limits_{x \in \mathcal{C}^{\nv + \nh}} \left\{ \boldsymbol t(\boldsymbol x)\dfrac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}}\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}\right\}.
\end{align*}
\hypertarget{instability}{%
\subsubsection{Instability}\label{instability}}

@@

In formulating a RBM model for a potentially large number of visibles
(i.e., as \(\nv \rightarrow \infty\)), we will say that the ratio
\eqref{eq:elpr} needs to stay bounded for a sequence of RBM models to be
stable. That is, we make the following convention.
\BeginKnitrBlock{definition}[S-unstable RBM]
\protect\hypertarget{def:unnamed-chunk-1}{}{\label{def:unnamed-chunk-1}
\iffalse (S-unstable RBM) \fi{} }Let
\(\boldsymbol \theta_\nv \in \mathbb{R}^{\nv + \nh + \nh*\nv}, \nv \ge 1\),
be an element of a sequence of RBM parameters where the number of
hiddens, \(\nh \equiv \nh(\nv) \ge 1\), can be an arbitrary function of
the number \(\nv\) of visibles. A RBM model formulation is
\emph{Schweinberger-unstable} or \emph{S-unstable} if \begin{align*}
\lim\limits_{\nv \rightarrow \infty} \frac{1}{\nv} \text{LREP}(\boldsymbol \theta_\nv) = \infty.
\end{align*}
\EndKnitrBlock{definition}

@@

\BeginKnitrBlock{proposition}
\protect\hypertarget{prp:instab1}{}{\label{prp:instab1} }Let \(c > 0\) and
let \(\text{LREP}(\boldsymbol \theta_\nv)\) be as in \eqref{eq:elpr} for
an integer \(\nv \ge 1\). If
\(\frac{1}{\nv}\text{LREP}(\boldsymbol \theta_\nv) > c\), then
\(\Delta(\boldsymbol \theta_\nv) > c\).
\EndKnitrBlock{proposition}

In other words, if the probability ratio \eqref{eq:elpr} is too large,
then a RBM model sequence will exhibit large probability shifts for very
small changes in data configurations (i.e., will exhibit instability).
Such instability can be a concern in a model for reasons similar to
those related to degeneracy: as outcomes vary over the sample space, the
geography of probabilities is extremely rugged, with deep pits following
sharp mountains. Recall the applied example of RBM models as a means to
classify images. For data as pixels in an image, the instability result
in Proposition \ref{prp:instab1} manifests itself as a one pixel change
in an image (one component of the visible vector) resulting in a large
shift in the probability, which in turn could result in a vastly
different classification of the image. Examples of this behavior have
been presented in Szegedy et al.
(\protect\hyperlink{ref-szegedy2013intriguing}{2013}) for other deep
learning models, in which a one pixel change in a test image results in
a wildly different classification.
